Higher-Order unification for free!

### 2024-05-03 paper/main.tex
minor
small update of proof: Fidelity with map-deduplication

\begin{proof}[Proof sketch]
  Let $X$ be a \Fo{}-variable appearing in two different subterms $s_1$ and $s_2$.
  In $s_1$ variable $X$ is applied to the list of terms $x_1\dots x_m$ while, in $s_2$,
  it is applied to $y_1\dots y_n$. Without loss of generality we take $m < n$.
  After \mapdeduplication we have two distinct \Ho{}-variables for $X$, say $A$ and $C$,
  related by a chain \chain of \linketa{}s of length $n - m$. We prove that
  if $A$ and $C$ are assigned respectively to the \wellb terms $\lambda x_1\dots
  x_p.t_1$ and $\lambda x_1\dots x_q.t_2$, then a unification failure occurs iff
  these terms fails to unify.\todo{a unification fails if they fail to unify is a tautology, to be clarified} By \progressetaleft, the assignment of $A$
  activates $p$ links of \chain. In particular, when $p \geq n$, then the entire
  chain collapses and unification is performed between $\lambda x_n\dots
  x_p.t_1$ and $\lambda x_1\dots x_q.t_2$. By \cref{th:fidelity-beta} and since
  we work with \wellb terms, this unification succeeds iff it does in \Fo.
  Otherwise if $p < n$ then we have to consider two cases. 1) If $t_1$ is a rigid
  term then $t_2$ must be the $\eta$-expansion of $t_1$, implying $q = 0$. If it
  is not the case, then \progressetaleft
  would eventually cause a failure. 2) If $t_1$ is a
  unification variable then \elpiIn{scope-check} guarantees that $t_2$ does not use
  $x_1\dots x_p$ as free variables, causing otherwise a unification failure.
  After this check, $m - n - p$ links of \chain remain suspended, as we expect:
  decompilation is charged to wake all links in \linkStore and eventually relate
  $t_1$ with $t_2$.
  %
  % Attenzione a questo caso:
  %    A =𝜂 λx.(B x)   e    x |- B x =𝜂 λx.(C x)           (from a map dedup, where A e C, sono stessa variabile con diff arities)
  %  se (C x) è instanziato a (f (D x)) allora non c'è nessun progress:
  %  il rhs del secondo link diventa (λx.(f (D x))) che è un maybe-eta
  % 
  % Nella nuova versione sottintendo questo caso: p = 0 e t_1 è flessibile
  %
  % Without loss of generality, we pose $m < n$. If $A^m$ is assigned to a term
  % $\lambda \vec{x}.t$ with $q$ being the length of $\vec{x}$. If $t$ is a rigid
  % term, then the chain of links between $A^m$ and $C^n$ is completely triggered
  % so that $C^n$ is assigned to $\lambda \vec{x}.t$ $\eta$-expanded $n-m$ times.
  % Note that in this case fidelity is guaranteed, since the term unified with $C^n$
  % sohuld also unify with $A^m$. Otherwise, if $t$ is a variable, the chain of
  % links is shrinked by $q$. Note that if this length is bigger then $n-m$
  % the full chain collapse. % and $A^m$ and $C^n$ are agian tiny coupled. 
  % Otherwise,
  % if $q < n-m$, it means that $C^n$ can always be assigned to any term starting
  % with at least $q$ lambdas. In this second case fidelity is also guaranteed,
  % due to the constraint of the number of lambda by which $C^n$ should start.
  %
  % not starting with $\lambda$, then the entrire chain of link is
  % immediately waken up, assigning $C^n$ to $x_1\dots x_{n-m} \vdash t \appsep
  % x_1\dots x_{n-m}$, which is the $t$ $\eta$-expanded $n-m$ times. Hence, $A^m$
  % and $C^n$ are tiny coupled. If $A^m$ is assigned to a term of the form
  % $\lambda \vec{x}.t$, then if $t$ is rigid, the chain of links between $A^m$
  % and $C^n$ completely triggered, is shrinked by the length of $\vec{x}$. Note
  % that if $t$ is rigid, by the previous sentence, the remi
  %
  % If $A^m$ is instantiated, 
  %
  % Whenever the two variables are assigned to the non-unifiable terms,
  % the chain of \linketa is entirely waken up refining little by little
  % the intermediate variables between $A^m$ and $C^n$ leading eventually to a failure. 
  % On the other
  % hand, if the two variables still remain unifiable after their assignement, 
  % then the activation of 
  % the corresponding \linketa just refines the chain, by reducing eventually
  % its length.
\end{proof}

@@

We considered mechanizing our results with Abella~\cite{gacek2008abella},
especially in the early phase of this work, but unfortunately we could not.
The main reason is that the logic of Abella does not let one quantify over
predicates nor use the cut operator. Eliminating cut or
higher order
combinators like \elpiIn{map} or \elpiIn{forall2} is surely
possible, but 
has a high cost in verbosity given how pervasively they are used.
Moreover, and more importantly, it creates a distance
between the verified and the actual code, which we found to
hinder our exploration. We did, however, establish a
test suite with about a hundred cases.
It is available at \githubUrl.

>>>>

\begin{proof}[Proof sketch]
  Let $X$ be a \Fo{}-variable appearing in two different subterms $s_1$ and $s_2$.
  In $s_1$ variable $X$ is applied to the list of terms $x_1\dots x_m$ while, in $s_2$,
  it is applied to $y_1\dots y_n$. Without loss of generality we take $m < n$.
  After \mapdeduplication we have two distinct \Ho{}-variables for $X$, say $A$ and $C$,
  related by a chain \chain of \linketa{}s of length $n - m$. We prove that
  if $A$ and $C$ are assigned respectively to the \wellb terms $\lambda x_1\dots
  x_p.t_1$ and $\lambda x_1\dots x_q.t_2$, then a failure occurs iff
  these terms do not unify. By \progressetaleft, the assignment of $A$
  activates $p$ links of \chain. In particular, when $p \geq n$, then the entire
  chain collapses and unification is performed between $\lambda x_n\dots
  x_p.t_1$ and $\lambda x_1\dots x_q.t_2$. By \cref{th:fidelity-beta} and since
  we work with \wellb terms, this unification succeeds iff it does in \Fo.
  Otherwise if $p < n$ then we have to consider two cases. 1) If $t_1$ is a rigid
  term then $t_2$ must be the $\eta$-expansion of $t_1$, implying $q = 0$. If it
  is not the case, then \progressetaleft
  would eventually cause a failure. 2) If $t_1$ is a
  unification variable then \elpiIn{scope-check} guarantees that $t_2$ does not use
  $x_1\dots x_p$ as free variables, causing otherwise a unification failure.
  After this check, $m - n - p$ links of \chain remain suspended, as we expect:
  decompilation is charged to wake all links in \linkStore and eventually relate
  $t_1$ with $t_2$.
  %
  % Attenzione a questo caso:
  %    A =𝜂 λx.(B x)   e    x |- B x =𝜂 λx.(C x)           (from a map dedup, where A e C, sono stessa variabile con diff arities)
  %  se (C x) è instanziato a (f (D x)) allora non c'è nessun progress:
  %  il rhs del secondo link diventa (λx.(f (D x))) che è un maybe-eta
  % 
  % Nella nuova versione sottintendo questo caso: p = 0 e t_1 è flessibile
  %
  % Without loss of generality, we pose $m < n$. If $A^m$ is assigned to a term
  % $\lambda \vec{x}.t$ with $q$ being the length of $\vec{x}$. If $t$ is a rigid
  % term, then the chain of links between $A^m$ and $C^n$ is completely triggered
  % so that $C^n$ is assigned to $\lambda \vec{x}.t$ $\eta$-expanded $n-m$ times.
  % Note that in this case fidelity is guaranteed, since the term unified with $C^n$
  % sohuld also unify with $A^m$. Otherwise, if $t$ is a variable, the chain of
  % links is shrinked by $q$. Note that if this length is bigger then $n-m$
  % the full chain collapse. % and $A^m$ and $C^n$ are agian tiny coupled. 
  % Otherwise,
  % if $q < n-m$, it means that $C^n$ can always be assigned to any term starting
  % with at least $q$ lambdas. In this second case fidelity is also guaranteed,
  % due to the constraint of the number of lambda by which $C^n$ should start.
  %
  % not starting with $\lambda$, then the entrire chain of link is
  % immediately waken up, assigning $C^n$ to $x_1\dots x_{n-m} \vdash t \appsep
  % x_1\dots x_{n-m}$, which is the $t$ $\eta$-expanded $n-m$ times. Hence, $A^m$
  % and $C^n$ are tiny coupled. If $A^m$ is assigned to a term of the form
  % $\lambda \vec{x}.t$, then if $t$ is rigid, the chain of links between $A^m$
  % and $C^n$ completely triggered, is shrinked by the length of $\vec{x}$. Note
  % that if $t$ is rigid, by the previous sentence, the remi
  %
  % If $A^m$ is instantiated, 
  %
  % Whenever the two variables are assigned to the non-unifiable terms,
  % the chain of \linketa is entirely waken up refining little by little
  % the intermediate variables between $A^m$ and $C^n$ leading eventually to a failure. 
  % On the other
  % hand, if the two variables still remain unifiable after their assignement, 
  % then the activation of 
  % the corresponding \linketa just refines the chain, by reducing eventually
  % its length.
\end{proof}

@@

We considered mechanizing our results with Abella~\cite{gacek2008abella},
especially in the early phase of this work, but unfortunately we could not.
The main reason is that the logic of Abella does not let one quantify over
predicates nor use the cut operator. Eliminating cut or
higher order
combinators like \elpiIn{map} or \elpiIn{forall2} is surely
possible, but 
has a high cost in verbosity given how pervasively they are used.
Moreover, and more importantly, it creates a distance
between the verified and the actual code, which we found to
hinder our exploration. However, we did establish a
test suite with about a hundred cases.
It is available at \githubUrl.
